---
title: "`SpatialData.plot`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('SpatialData.plot')`"
author: 
  - name: Helena Lucia Crowell
  - name: Charlotte Soneson
  - name: Michael Stadler
  - name: Daria Lazic
  - name: Samuel Gunz
  - name: Rafael Irizarry
  - name: Hugo Gruson
  - name: Gabriel Grajeda
  - name: Martin Emons
output:
  BiocStyle::html_document
self-contained: true
vignette: |
  %\VignetteIndexEntry{SpatialData.plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
```

```{r load-libs}
library(ggplot2)
library(patchwork)
library(ggnewscale)
library(SpatialData)
library(arrow)
library(SpatialData.plot)
library(SingleCellExperiment)
library(sosta)
library(sf)
```

# Introduction

This project is concerned with enhancing the interoperability between `python` and `R`/ `Bicondocutor`. As a use-case we use a `SpatialData` object of breast cancer from Janesick et al. acquired by 10x Genomics Xenium. We load the `SpatialData` object in `python` and run a `python` method on it called `SegTraQ`. 

## Run `python` module `SegTraQ`

```{python run SegTraQ, eval = FALSE}
import segtraq
import spatialdata as sd

sdata = sd.read_zarr("xenium.zarr")

st = segtraq.SegTraQ(
    sdata, images_key="image", tables_centroid_x_key="x_centroid", tables_centroid_y_key="y_centroid"
)

st.run_baseline()

st.sdata.write("xenium_with_qc.zarr", overwrite=True)
```

## Run `R` module `spacexr`

With this object we then load the `SpatialData` object in R. We want to compute further metrics in `R` namely the `RCTD` algortihm implemented in `spacexr`.

```{r run RCTD, eval = FALSE}
library(spacexr)
library(SummarizedExperiment)
library(anndataR)

# Read scRNA-seq annotation
sc_annotation_filepath <- "BC_scRNAseq_Janesick.h5ad"
sc_annotation <- anndataR::read_h5ad(sc_annotation_filepath, as = "SingleCellExperiment")


# Read spatial data
zarr_filepath <- "xenium_with_qc.zarr"
spatial_data_filepath <- paste0(zarr_filepath, "/tables/table")
spatial_data <- read_zarr(spatial_data_filepath, as = "SingleCellExperiment")

# Run RCTD
rctd_config <- createRctd(spatial_data, sc_annotation, cell_type_col="Annotation", pixel_count_min=0, UMI_min=0)
rctd_results <- runRctd(rctd_config, rctd_mode="full", max_cores=7, doublet_threshold=0)

# Shannon entropy function: H = -sum(p * log(p))
entropy <- function(col) {
  p <- col[col > 0]  # avoid log(0)
  max(0, -sum(p * log(p)))   # using natural log, clip to zero for numerical errors
}

# Compute per-cell entropy
cell_entropies <- apply(
  as.matrix(assay(rctd_results, "weights_full")),
  MARGIN = 2,
  FUN = entropy
)

# Get numeric positions of RCTD cells within spatial_data
idx <- match(names(cell_entropies), colnames(spatial_data))

# Drop any NAs (cells in RCTD not found in spatial_data)
valid <- !is.na(idx)

colData(spatial_data)$rctd_entropy <- rep(NA_real_, ncol(spatial_data))
colData(spatial_data)$rctd_entropy[idx[valid]] <- cell_entropies[valid]

colData(spatial_data)$rctd_singlet_scores <- rep(NA_real_, ncol(spatial_data))
colData(spatial_data)$rctd_singlet_scores[idx[valid]] <- rctd_results$singlet_score[valid]

adata <- as_AnnData(spatial_data)
adata$write_h5ad(
  "./updated_sdata_xenium_with_rctd.h5ad",
  mode="w"
)
```

we then merge the results into our `SpatialData` `SingleCellExperiment` object

```{r read SpatialData}
(x <- readSpatialData("xenium_with_qc.zarr", anndataR=TRUE))
```

```{r, read spacexrResults}
h5ad_file <- "updated_adata_with_rctd.h5ad"
sce_obj <- anndataR::read_h5ad(h5ad_file, as = "SingleCellExperiment")
```

```{r}
library(dplyr)
colData(tables(x)[[1]])$cell_id <- int_colData(tables(x)[[1]])$cell_id

df1 <- colData(tables(x)[[1]]) |> as.data.frame()
df2 <- colData(sce_obj)[c("cell_id","rctd_entropy", "rctd_singlet_scores")] |> as.data.frame()

merged <- df1 |>
  dplyr::left_join(df2, by = "cell_id")

colData(tables(x)[[1]]) <- as(merged, "DataFrame")
```

## Plot results with `R` package `SpatialData.plot`

### 

In order to plot this we use the `SpatialData.plot` library implemented to interface with the `SpatialData` object

```{r plotPoint, eval=TRUE}
y <- x$points$transcripts
#filter interesting genes
y <- y |> dplyr::filter(feature_name  %in% c("EPCAM", "LUM"))
# demo. viz. capabilities
plotSpatialData() + 
    plotShape(x, i = "cell_boundaries", fill = "transferred_celltype_plot", alpha = 1, linewidth = 0) + 
    #scale_fill_viridis_c() +
    scale_fill_manual(values=unname(pals::trubetskoy())) +
    new_scale("fill") + 
    plotShape(x, i = "nucleus_boundaries", color = "white", alpha = 0, linewidth = 0) + 
    new_scale("fill") + 
    plotPoint(y, colour = "feature_name", size = 0.3)

  plotSpatialData() + 
    plotShape(x, i = "cell_boundaries", fill = "rctd_entropy", alpha = 1, linewidth = 0) + 
    scale_fill_viridis_c() +
    #scale_fill_manual(values=unname(pals::trubetskoy())) +
    new_scale("fill") + 
    plotShape(x, i = "nucleus_boundaries", color = "white", alpha = 0, linewidth = 0) + 
    new_scale("fill") + 
    plotPoint(y, colour = "feature_name", size = 0.3)
```

```{r}
bb_xmin = 800
bb_ymin = 1150
bb_w = 200
bb_h = 300
bb_xmax = bb_xmin + bb_w
bb_ymax = bb_ymin + bb_h
bbox <- list(xmin = bb_xmin, xmax = bb_xmax, ymin = bb_ymin, ymax = bb_ymax)
shape(x, "a") <- query(shape(x, "cell_boundaries"),bbox)
shape(x, "b") <- query(shape(x, "nucleus_boundaries"),bbox)
point(x, "c") <- query(point(x),bbox)
#tables(x, "d") <- query(tables(x),bbox)

t = tables(x)[[1]]
t@int_colData$region <- "a"
t@int_metadata$spatialdata_attrs$region <- "a"

i = int_colData(t)$cell_id %in% shape(x, "a")$cell_id
t = t[, i]

table(x, "d") <- t

#filter interesting genes
y <- point(x, "c") |> dplyr::filter(feature_name  %in% c("EPCAM", "LUM"))
# demo. viz. capabilities
p <- plotSpatialData() + 
    plotShape(x, i = "a", fill = "transferred_celltype_plot", alpha = 1, linewidth = 0) +
   scale_fill_manual(values=unname(pals::trubetskoy()))

q <- plotSpatialData() + 
    plotShape(x, i = "a", fill = "F1_purity", alpha = 1, linewidth = 0) + 
    scale_fill_viridis_c() +
    new_scale("fill") + 
    plotShape(x, i = "b", colour = "white", alpha = 0, linewidth = 0) + 
    new_scale("colour") + 
    plotPoint(y, colour = "feature_name", size = 0.3)

r <- plotSpatialData() + 
    plotShape(x, i = "a", fill = "rctd_entropy", alpha = 1, linewidth = 0) + 
    scale_fill_viridis_c() +
    new_scale("fill") + 
    plotShape(x, i = "b", colour = "white", alpha = 0, linewidth = 0) + 
    new_scale("colour") + 
    plotPoint(y, colour = "feature_name", size = 0.3)
```

```{r}
library(patchwork)
p
q
r
```

# Reconstruction,  plotting anatomical regions

Here we will use `sosta` to reconstruct the DCIS2 regions, select them and later plot them. We write a little helper function to directly call `sosta` on the `SpatialData` object.

```{r}
sostaPolygonShape <- function(x, # SD class
                              shapeName, # name of the shape layer
                              marks, # mark column of interest
                              markSelect, # selected mark for reconstruction
                              dim = 500, # reconstruction dim
                              ... # other sosta parameters
                              ){
 # create the centroids from a shape layer 
 crd <- st_centroid(shape(x, shapeName)$geometry) |>
  st_coordinates() 
 
 # select the column data with any relevant metadata
 cd <- getTable(x, shapeName) |>
  colData()
 # combine the crd and the column data
 pp <- .df2ppp(cbind(crd, cd),
               xName = colnames(crd)[[1]],
               yName = colnames(crd)[[2]],
               marks = marks)
 # reconstruct the poly
 poly <- reconstructShapeDensity(pp, markSelect, dim = dim, ...)
 return(poly)
}
```

We will run `sosta` to reconstruct the DCIS2 regions.

```{r}
res <- sostaPolygonShape(
 x = x,
 shapeName = "cell_boundaries",
 marks = "transferred_celltype_plot",
 markSelect = "DCIS2"
)
```

For illustration we will select the largest to polygons.

```{r}
resSel <- res[order(st_area(res), decreasing = TRUE)[1:2],]
```

Now we do a shape query. This accepts also multiple polygons.

```{r}
shape(x, "a") <- query(shape(x, "cell_boundaries"), resSel)
shape(x, "b") <- query(shape(x, "nucleus_boundaries"), resSel)
point(x, "c") <- query(point(x), resSel)
```

```{r}
t = tables(x)[[1]]
t@int_colData$region <- "a"
t@int_metadata$spatialdata_attrs$region <- "a"

i = int_colData(t)$cell_id %in% shape(x, "a")$cell_id
t = t[, i]

table(x, "d") <- t

#filter interesting genes
y <- point(x, "c") |> dplyr::filter(feature_name  %in% c("EPCAM", "LUM"))
```

For now the `sosta` output is not yet stored in the `SpatialData` object. We will plot it using `geom_sf()`.

```{r}
# demo. viz. capabilities
plotSpatialData() + 
 plotShape(x, i = "a", fill = "transferred_celltype_plot", alpha = 1, linewidth = 0) + 
 scale_fill_manual(values=unname(pals::trubetskoy())) +
 new_scale("fill") + 
 plotShape(x, i = "b", colour = "white", alpha = 0, linewidth = 0) + 
 new_scale("colour") + 
 plotPoint(y, colour = "feature_name", size = 0.1) +
 geom_sf(data = resSel, fill = NA, inherit.aes = FALSE, color = "navy",
         linewidth = 0.75) +
 labs(title = "Multipolygon query")
```

# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
